Analysis of finite-volume transport schemes on cubed-sphere grids and an accurate scheme for divergent winds

### 2024-05-27 paper.tex
updated title: advection -> transport
updated contact
minor correction
update acknowledgments

\title{Analysis of finite-volume advection schemes on cubed-sphere grids and an accurate scheme for divergent winds}

@@

\author[1]{Luan F. Santos\corref{cor1}}
\ead{luan.santos@usp.br}

@@

In FV3, it is assumed that the metric $\sqrt{\mathfrak{g}}$ appearing in Equation \eqref{numerical-flux} is constant over the integration domain, specifically equal to $\sqrt{\mathfrak{g}}_{i+\frac{1}{2}}$, hence only $q$ needs to be reconstructed instead of $\sqrt{\mathfrak{g}}q$.
This implies the following approximation in Equation \eqref{1d-avflux-approx}:
\begin{equation}
	\label{1d-avflux-approx}
	\int^{x_{i+\frac{1}{2}}}_{ x_{i+\frac{1}{2}}^d(t^n,t^{n+1})} (\sqrt{\mathfrak{g}}q)(x,t^n)\,dx = 
	\sqrt{\mathfrak{g}}_{i+\frac{1}{2}}
	\int^{x_{i+\frac{1}{2}}}_{ x_{i+\frac{1}{2}}^d(t^n,t^{n+1})} q(x,t^n)\,dx + \mathcal{O}(\Delta x).
\end{equation}
The scheme proposed in this work, on the other hand, considers the reconstruction of $\sqrt{\mathfrak{g}}q$, avoiding the approximation in Equation \eqref{1d-avflux-approx}.
The reconstruction employed here for both the FV3 and LT2 schemes uses the PPM scheme from \cite{colella:1984,carpenter:1990}.
We introduce the grid function $q^M$ expressed as:
\begin{align}
	\label{ppm_q_m}
    q^M = 
	\begin{cases}
        q^n,
		\quad &\text{if }{M=\text{FV3}},\\
        \sqrt{\mathfrak{g}}q^n,
		\quad &\text{if }{M=\text{LT2}}.
	\end{cases}
\end{align}
Following \cite{harris:2021}, the PPM reconstruction on the $i$-th cell for a grid function $q^M$ may be expressed as:
\begin{align}
	\label{ppm-formula}
	\widetilde{q^M}(x) =
	\widetilde{q^M_i}(x) =
	a_1+ a_2
	\bigg(\frac{x-x_{i-\frac{1}{2}}}{\Delta x}\bigg)+
	a_3
	\bigg(\frac{x-x_{i-\frac{1}{2}}}{\Delta x}\bigg)^2,
\end{align}
for $x \in [x_{i-\frac{1}{2}},x_{i+\frac{1}{2}}]$, where
\begin{align}
	a_1 = q_{L,i}^M , \quad
	a_2 = 4b_{L,i}^M +2b_{R,i}^M, \quad
	a_3 = 3(b_{L,i}^M +b_{R,i}^M ) , 
\end{align}
where the following perturbation values of  ${q_i^M}$ named by \cite{harris:2021} are introduced
\begin{align}
	b_{L,i}^M = q_{L,i}^M-{q_i^M}, \quad 
	b_{R,i}^M = q_{R,i}^M-{q_i^M},
\end{align}
and ${q_i^M}$
are the average or centroid values at the $i$-th cell.
It is easy to see that integrating Equation \eqref{ppm-formula} on $x \in [x_{i-\frac{1}{2}},x_{i+\frac{1}{2}}]$ gives $q^M_i\Delta x$, therefore the PPM reconstructions preserves the mass on each cell. It is also easy to see that:
\begin{align}
	\lim_{x \to x_{i-\frac{1}{2}}^+}{q^M}(x) = q_{L,i}^M,\quad
	\lim_{x \to x_{i+\frac{1}{2}}^-}{q^M}(x) = q_{R,i}^M.
\end{align}
Therefore, the values $q_{L,i}^M$ and $q_{R,i}^M$ should approximate ${q_{i-\frac{1}{2}}^M}$ and ${q_{i+\frac{1}{2}}^M}$, respectively.
Furthermore, we consider:
\begin{align}
	\label{phi_L}
	q_{R,i}^M  =
	\begin{cases}
		q_{R,i},
		\quad &\text{if }{M=\text{FV3}},\\
		\sqrt{\mathfrak{g}}_{i+\frac{1}{2}} q_{R,i},
		\quad &\text{if }{M=\text{LT2}},
	\end{cases}
\end{align}
and similarly to $q_{L,i}^M$, where 
$q_{L,i}$ and $q_{R,i}$ should approximate ${q_{i-\frac{1}{2}}^n}$ and ${q_{i+\frac{1}{2}}^n}$, respectively.
One possible way of doing this approximation is by employing a reconstruction method based on primitive functions \cite[Chapter 17]{leveque:1990}.
For instance, \cite{colella:1984} uses
\begin{equation}
	\label{unlim-ppm}
	q_{R,i} = \frac{7}{12} \bigg( q_{i+1}^n + q_{i}^n \bigg) - \frac{1}{12} \bigg(q_{i+2}^n +q_{i-1}^n\bigg),
\end{equation}
and $q_{L,i}=q_{R,i-1}$ for the unlimited PPM reconstruction.
This formula is fourth-order accurate if the exact average values are used and second-order accurate if the centroid values are used.
This scheme is refereed to as UNLIM.
This work also considers the monotonic scheme outlined in \cite[Appendix B]{lin:2004}, where the values of $q_{L,i}$ and $q_{R,i}$ are determined by equations B3 and B4 in the same appendix.
This scheme is referred to as MONO.
Finally, both UNLIM and MONO schemes require $\nu=3$ layers of ghost cells.

@@

\section*{Acknowledgments}
We acknowledge the financial support provided by the São Paulo Research Foundation (FAPESP) under the grants 2020/10280-4 and 2021/06176-0.

>>>>

\title{Analysis of finite-volume transport schemes on cubed-sphere grids and an accurate scheme for divergent winds}

@@

\author[1]{Luan F. Santos\corref{cor1}}
\ead{luanfsantos14@gmail.com}

@@

In FV3, it is assumed that the metric $\sqrt{\mathfrak{g}}$ appearing in Equation \eqref{numerical-flux} is constant over the integration domain, specifically equal to $\sqrt{\mathfrak{g}}_{i+\frac{1}{2}}$, hence only $q$ needs to be reconstructed instead of $\sqrt{\mathfrak{g}}q$.
This implies the following approximation in Equation \eqref{1d-avflux-approx}:
\begin{equation}
	\label{1d-avflux-approx}
	\int^{x_{i+\frac{1}{2}}}_{ x_{i+\frac{1}{2}}^d(t^n,t^{n+1})} (\sqrt{\mathfrak{g}}q)(x,t^n)\,dx = 
	\sqrt{\mathfrak{g}}_{i+\frac{1}{2}}
	\int^{x_{i+\frac{1}{2}}}_{ x_{i+\frac{1}{2}}^d(t^n,t^{n+1})} q(x,t^n)\,dx + \mathcal{O}(\Delta x).
\end{equation}
The scheme proposed in this work, on the other hand, considers the reconstruction of $\sqrt{\mathfrak{g}}q$, avoiding the approximation in Equation \eqref{1d-avflux-approx}.
The reconstruction employed here for both the FV3 and LT2 schemes uses the PPM scheme from \cite{colella:1984,carpenter:1990}.
We introduce the grid function $q^M$ expressed as:
\begin{align}
	\label{ppm_q_m}
    q^M = 
	\begin{cases}
        q^n,
		\quad &\text{if }{M=\text{FV3}},\\
        \sqrt{\mathfrak{g}}q^n,
		\quad &\text{if }{M=\text{LT2}}.
	\end{cases}
\end{align}
Following \cite{harris:2021}, the PPM reconstruction on the $i$-th cell for a grid function $q^M$ may be expressed as:
\begin{align}
	\label{ppm-formula}
	\widetilde{q^M}(x) =
	\widetilde{q^M_i}(x) =
	a_1+ a_2
	\bigg(\frac{x-x_{i-\frac{1}{2}}}{\Delta x}\bigg)+
	a_3
	\bigg(\frac{x-x_{i-\frac{1}{2}}}{\Delta x}\bigg)^2,
\end{align}
for $x \in [x_{i-\frac{1}{2}},x_{i+\frac{1}{2}}]$, where
\begin{align}
	a_1 = q_{L,i}^M , \quad
	a_2 = -(4b_{L,i}^M +2b_{R,i}^M), \quad
	a_3 = 3(b_{L,i}^M +b_{R,i}^M ) , 
\end{align}
where the following perturbation values of  ${q_i^M}$ named by \cite{harris:2021} are introduced
\begin{align}
	b_{L,i}^M = q_{L,i}^M-{q_i^M}, \quad 
	b_{R,i}^M = q_{R,i}^M-{q_i^M},
\end{align}
and ${q_i^M}$
are the average or centroid values at the $i$-th cell.
It is easy to see that integrating Equation \eqref{ppm-formula} on $x \in [x_{i-\frac{1}{2}},x_{i+\frac{1}{2}}]$ gives $q^M_i\Delta x$, therefore the PPM reconstructions preserves the mass on each cell. It is also easy to see that:
\begin{align}
	\lim_{x \to x_{i-\frac{1}{2}}^+}{q^M}(x) = q_{L,i}^M,\quad
	\lim_{x \to x_{i+\frac{1}{2}}^-}{q^M}(x) = q_{R,i}^M.
\end{align}
Therefore, the values $q_{L,i}^M$ and $q_{R,i}^M$ should approximate ${q_{i-\frac{1}{2}}^M}$ and ${q_{i+\frac{1}{2}}^M}$, respectively.
Furthermore, we consider:
\begin{align}
	\label{phi_L}
	q_{R,i}^M  =
	\begin{cases}
		q_{R,i},
		\quad &\text{if }{M=\text{FV3}},\\
		\sqrt{\mathfrak{g}}_{i+\frac{1}{2}} q_{R,i},
		\quad &\text{if }{M=\text{LT2}},
	\end{cases}
\end{align}
and similarly to $q_{L,i}^M$, where 
$q_{L,i}$ and $q_{R,i}$ should approximate ${q_{i-\frac{1}{2}}^n}$ and ${q_{i+\frac{1}{2}}^n}$, respectively.
One possible way of doing this approximation is by employing a reconstruction method based on primitive functions \cite[Chapter 17]{leveque:1990}.
For instance, \cite{colella:1984} uses
\begin{equation}
	\label{unlim-ppm}
	q_{R,i} = \frac{7}{12} \bigg( q_{i+1}^n + q_{i}^n \bigg) - \frac{1}{12} \bigg(q_{i+2}^n +q_{i-1}^n\bigg),
\end{equation}
and $q_{L,i}=q_{R,i-1}$ for the unlimited PPM reconstruction.
This formula is fourth-order accurate if the exact average values are used and second-order accurate if the centroid values are used.
This scheme is refereed to as UNLIM.
This work also considers the monotonic scheme outlined in \cite[Appendix B]{lin:2004}, where the values of $q_{L,i}$ and $q_{R,i}$ are determined by equations B3 and B4 in the same appendix.
This scheme is referred to as MONO.
Finally, both UNLIM and MONO schemes require $\nu=3$ layers of ghost cells.

@@

\section*{Acknowledgments}
We acknowledge the financial support provided by the São Paulo Research Foundation (FAPESP) under the grants 2020/10280-4 and 2021/06176-0 and The National Council for Scientific and Technological Development (CNPq), grant 303436/2022-0. 
This study was financed in part by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior - Brasil (CAPES) - Finance Code 001.
